### 2014-10-30 TransitionsUsefulThird.tex
stuff

\section{Labeled Transition Systems}\label{sec:lts}

We make extensive use of the general theory of labeled transition
systems, a semantics approach especially relevant to communicating
concurrent systems.  As we are formalizing processors for
Turing-complete machine languages, it is challenging to prove that a
system preserves almost any aspect of processor behavior from a model
like SC.  To focus our theorems, we pick the time-honored property of
\emph{termination}.  An optimized system should terminate or diverge
iff the reference system could also terminate or diverge,
respectively.  Our basic definitions of transition systems build in
special treatment of halting, so that we need not mention it
explicitly in most contexts to come.

\begin{defn}
A \textbf{labeled transition system (LTS)} is a ternary relation, over
$\mathcal S^H \times \mathcal L^\epsilon \times \mathcal S^H$, for some sets
$\mathcal S$ of states and $\mathcal L$ of labels. We usually do not mention
these sets explicitly, as they tend to be clear from context. We write
$X^\epsilon$ for lifting of a set $X$ to have an extra ``empty'' element
$\epsilon$ (like an \texttt{option} type in ML). We write $X^H$ for lifting of
a set $X$ to have an extra ``halt'' element $H$. We also implicitly consider
each LTS to be associated with an \emph{initial} state in $\mathcal S$.
\end{defn}

For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
intuition is that $A$ is one process within a concurrent system. The
label $\ell$ from set $\mathcal L$ of labels is produced when $A$
participates in some IO exchange with another process; otherwise it is
an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
omit labels for $\epsilon$ steps.  Note that the IO exchange can be
input, output, or both, and its interpretation is dependent on the
context.

\subsection{Basic Constructions on LTSes}

From an LTS representing single-step system evolution, we often want to build
an LTS capturing arbitrary-length evolutions.

\begin{defn}
The \textbf{transitive-reflexive closure} of $A$, written $A^{*}$, is a derived
LTS.  Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the
states of $A^{*}$ are $\mathcal S$, and the labels are $\mathcal L^{*}$, or
\emph{sequences} of labels from the original system.  $A^{*}$ steps from $s$ to
$s'$ when there exist zero or more transitions in $A$ that move from $s$ to
$s'$.  The label of this transition is the \emph{concatenation} of all labels
generated in $A$, where the empty or ``silent'' label $\epsilon$ is treated as
an identity element for concatenation.
\end{defn}

We also want to compose $n$ copies of an LTS together, with no explicit
communication between them.  We apply this construction later to lift a
single-CPU system to a multi-CPU system.

\begin{defn}
The \textbf{$n$-repetition} of $A$, written $\smulti{n}{A}$, is a derived LTS.
Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the states of
$\smulti{n}{A}$ are $\mathcal S^n$, and the labels are $[1, n] \times \mathcal
L$, or pairs that \emph{tag} labels with which component system generated them.
These labels are generated only when the component system generates a label.
The whole system halts whenever one of the components halts.  We define the
transition relation with the following inference rules. We treat $n$-tuples as
functions keyed on numeric positions, and we write $\theta[i := v]$ for
updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and
we treat any label $(i, \epsilon)$ as $\epsilon$.
$$\inference[i]
{\theta(i) = a & \io{$A$}{a}{a'}{\ell} & a' \neq H}
{\io{$\smulti{n}{A}$}{\theta}{\theta[i:=a']}{i,\ell}}
\quad \inference[H]
{\io{$A$}{\theta(i)}{H}{\ell}}
{\io{$\smulti{n}{A}$}{\theta}{H}{i,\ell}}
$$
\end{defn}

Eventually, we need processes to be able to communicate with each
other, which we formalize via the $+$ composition operator. It
connects same-label transitions in the two systems, treating the label
as a cooperative communication event that may now be hidden from the
outside world, as an $\epsilon$ label.

\begin{defn}\label{plus}
Where $A$ and $B$ are two LTSes sharing labels set $\mathcal L$,
and with state sets $\mathcal S_A$ and
$\mathcal S_B$ respectively, the \textbf{communicating composition} $A + B$ is
a new LTS with states $\mathcal S_A \times \mathcal S_B$ and an empty label set,
defined as follows:
\vspace{-.2in}\begin{center}
$$\inference[$A$]
{\io{$A$}{a}{a'}{} & a' \neq H}
{\io{$A + B$}{a, b}{a', b}{}}
\quad \inference[$B$]
{\io{$B$}{b}{b'}{} & b' \neq H}
{\io{$A + B$}{a, b}{a, b'}{}}$$
$$\inference[H$_A$]
{\io{$A$}{a}{H}{}}
{\io{$A + B$}{a, b}{H}{}}
\quad \inference[H$_B$]
{\io{$B$}{b}{H}{}}
{\io{$A + B$}{a, b}{H}{}}$$
$$\inference[Join]
{\io{$A$}{a}{a'}{\ell} & \io{$B$}{b}{b'}{\ell} & a', b' \neq H}
{\io{$A + B$}{a, b}{a', b'}{}}$$
\end{center}
\end{defn}

\subsection{Refinement Between LTSes}

We need a notion of when one LTS ``implements'' another.  Intuitively,
transition labels and halting are all that the outside world
can observe. Two systems that produce identical labels and termination behavior
under all circumstances can be considered as safe substitutes for
one another. We will only need an asymmetrical notion of compatibility:

\begin{defn}\label{refines}
For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
be a function that is able to replace labels with alternative
labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
set $\mathcal L$. We say that
\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
\begin{multline*}
\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
(s_A = H \Leftrightarrow s_B = H)
\end{multline*}
Each label in the trace is replaced by the mapping of $f$ on it, and
labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
to denote that multi-label version when applied to $\eta$.
\end{defn}

As a shorthand, we write $A \sqsubseteq B$ for $A \sqsubseteq_\mathsf{id} B$,
for $\mathsf{id}$ an identity function, forcing traces in the two systems to
match exactly. Under this notion of identical traces, we say that $A$
is sound w.r.t. $B$.

\subsection{A Few Useful Lemmas}

The reflexivity and transitivity properties of refinement are crucial to help us
structure modular proofs.

\begin{theorem}\label{transitive}
$\sqsubseteq$ is reflexive and transitive.
\end{theorem}

Another crucial proof step is to lift a single-process result to the $n$-way
replication of that process.

\begin{theorem}\label{liftn}
If $A \sqsubseteq_f B$, then $\smulti{n}{A} \sqsubseteq_{f^n} \smulti{n}{B}$, where
$f^n$ is $f$ lifted appropriately to deal with indices ($f^n(i, \ell) =
(i, \ell')$ when $f(\ell) = \ell'$, and $f^n(i, \ell) = \epsilon$ when
$f(\ell) = \epsilon$).
\end{theorem}

We will also need a result applying to communicating compositions.

\begin{theorem}\label{liftplus}
If $A \sqsubseteq_f A'$ and $B \sqsubseteq_f B'$, then $A + B \sqsubseteq A' +
B'$. In other words, if systems $A$ and $B$ are individually simulated by $A'$ and
$B'$ on identical alteration of the traces, then the composed
system $A+B$ will be sound with respect to $A'+B'$.
\end{theorem}

%\begin{figure}
%\small
%\centering
%\begin{boxedminipage}[c]{.48\textwidth}
%%\centering
%\inference
%[Halt]
%{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
%{\io{SC}{\theta,m}{H}{}}
%
%\inference
%[NonMem]
%{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
%{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
%
%\inference
%[Load]
%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
%
%\inference
%[Store]
%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
%\end{boxedminipage}
%\caption{LTS for sequential consistency with $n$ simple processors}
%\label{Ref}
%\end{figure}
%
\section{Decomposing a Multi-Processor System}
\label{sec:store-atomicity}

\begin{figure}
\centering
\includegraphics[scale=.43]{zoom}
\caption{Components of a multi-processors system}
\label{zoom}
\end{figure}

Most any conventional multi-processor system can be divided into 3 components as shown in
Figure \ref{zoom}. 
We show the top-level system design at top right, and we zoom in on boxes standing for the
memory system and a single processor ($P_i$).
The processor component $P_i$ can be implemented in a
variety of ways, from literal execution of instructions in program order, to complex speculation
with instructions run simultaneously to exploit parallelism. The memory component is normally implemented
using a hierarchy of caches, in order to increase the performance of the
overall system, since the latency of accessing a memory directly is very slow
compared to accessing a much smaller cache. The component between each
processor and the global memory sub-system contains local buffers, $B_i$, specific to
the processor $P_i$ they are connected to.

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Load]
{}
{\io{$M_m$}{m}{m}{p, (\ld\req,a),(\ld\resp,m(a), \_)}}

\inference
[Store]
{}
{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}

\end{boxedminipage}
\caption{LTS for a simple memory}
\label{$M_m$}
\end{figure}

Popular instruction set architectures do not guarantee sequential
consistency.  That is, their behavior makes it unsound to imagine that
the different processors of a shared-memory system are running
according to simple interleaving, where each instruction executes
atomically.  However, we want to emphasize that, in every weak-memory
system we are aware of, \emph{the main memory still exposes atomic
  loads and stores}!  The weaker semantics arises only because of (1)
reordering of memory instructions by the core $P_i$ and/or (2) the
properties of the local buffers $B_i$ connected to each processor
$P_i$.  For instance, if $B_i$ were a store buffer, into which the
committed stores from an in-order processor $P_i$ are enqueued, and if
any ``subsequent'' load from the same processor $P_i$ reads the latest
value from the store buffer, then one can establish that such a
multi-processor system implements the TSO~\cite{x86tsocacm10} memory
model.

So, we focus on this opportunity to simplify proof decomposition.  We
will prove that our main memory component satisfies an intuitive
\emph{store atomicity} property.  Despite the potential red herring of
the word ``atomicity,'' this specification is appropriate even for
implementation of weaker memory models than SC.  Store atomicity can
be understood via the operational semantics of Figure \ref{$M_m$},
describing an LTS that receives load and store requests ($\ld$ and
$\st$) from processors and sends back load responses ($\ld\resp$);
$p$ is the processor which sends and receives the request.

The memory component is composed of a \emph{hierarchy of caches}, where each
cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
do our best to satisfy all memory requests, to avoid the latency of a
round-trip with main memory.  With an L1 cache miss, we may still find a hit in
a parent cache below the level of main memory, realizing a smaller but still
significant speedup.  Therefore it is the responsibility of the hierarchy of
caches, which forms the memory sub-component, to implement the store atomicity
property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
cache-coherence protocol is to establish this invariant for the memory
sub-component.  Concretely, we have verified a \emph{directory-based} protocol
for coordinating an arbitrary tree of caches, where each node stores a
conservative approximation of its children's states.

As an instance of the above decomposition, we will prove that a multi-processor
system with no local buffering in between the processor and the memory
components indeed implements sequential consistency. We implement a highly
speculative processor that executes instructions and issues loads out of
order, but commits instructions (once some ``verification'' is done) in order.

The processor itself can be decomposed into several components. In the zoomed-in
version of Figure \ref{zoom}, we show a highly speculative, out-of-order
issue processor. We have the normal architectural state, such as values of
registers.  Our proofs are generic over \emph{a family of instruction set
architectures}, with parameters for opcode sets and functions for executing
opcodes and decoding them from memory.  Other key components are a \emph{branch
predictor}, which guesses at the control-flow path that a processor will
follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
decides which instructions along that path to try executing ahead of schedule.
Our proofs apply to an arbitrary branch predictor, and they work for any
reorder buffer satisfying a simple semantic condition.

Figure~\ref{proofs} gives the overall proof structure that we employ to verify
the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
represents the speculative out-of-order processor, $M_c$ the cache memory,
$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
instruction atomically, thus implementing sequential consistency.

Our current framework establishes theorems of the form ``if system $A$ has a run
with some particular observable behavior, then system $B$ also has a run with
the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
Other important properties, such as \emph{deadlock freedom} for $A$ (which
might get stuck without producing any useful behavior), we leave for future
work.

\begin{figure}
\includegraphics[scale=.45]{proofs}
\caption{Overall proof structure}
\label{proofs}
\end{figure}

\section{Specifying Sequential Consistency}\label{sec:sc}

Our final theorem in this paper establishes that a particular complex hardware
system implements sequential consistency (SC) properly.  We state the theorem
in terms of the trace refinement relation $\sqsubseteq$.
Therefore, we need to define an LTS embodying SC.  The simpler this
system, the better.  We do not need to worry about its performance properties,
since we will prove that an optimized system remains faithful to it.

\begin{figure}
\small
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Halt]
{\dec(s,\pc) = H}
{\io{P$_\text{ref}$}{s,\pc}{H}{}}

\inference
[NM]
{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (s', \pc')}
{\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}

\inference
[Ld]
{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s', \pc')}
{\io{P$_\text{ref}$}{s, \pc}
{s', \pc'}{(\ld\req, a), (\ld\resp, v), \Real}}

\inference
[St]
{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s', \pc')}
{\io{P$_\text{ref}$}{s, \pc}
{s', \pc'}{(\st, a, v)}}
\end{boxedminipage}

\caption{LTS for a single reference processor (P$_{\text{ref}}$)}
\label{Pref$}
\end{figure}

Figure \ref{Pref$} shows the LTS for a single processor that will be used as a
reference processor.  The definition is parametrized over details of an
\emph{instruction set architecture} (ISA).  In particular, the ISA gives us
some domains of architectural states $s$ (e.g., register files) and of program
counters $\pc$.  A function $\dec(s, \pc)$ figures out which instruction $\pc$
references in the current state, returning the instruction's ``decoded'' form.
A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
instruction, returning the new state $s'$ and the next program counter $\pc'$.

The legal instruction forms, which are outputs of $\dec$, are $(\nm, x)$, for
an operation not accessing memory; $(\ld, x, a)$, for a memory load from
address $a$; $(\st, a, v)$, for a memory store of value $v$ to address $a$; and
$H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
$x$ above represents the rest of the instruction, including the opcode,
registers, constants, \etc{}

The label generated for a load contains the request sent to the
memory system, and the response obtained, as well as a label suffix $\Real$
which denotes that the load is non-speculative (We will later discuss the
alternative, $\Spec$ which arises in the load labels for our optimized
speculative processor).

The legal inputs to $\exec$ encode both a decoded instruction and any relevant
responses from the memory system.  These inputs are $(\nm, x)$ and $\st$, which
need no extra input from the memory; and $(\ld, x, v)$, where $v$ gives the
contents of the requested memory cell.

We can define sequential consistency using the simple memory from Figure
\ref{$M_m$} as follows, for $n$ processors:
\begin{defn}
$SC \triangleq P^n_\text{ref} + M_m$
\label{sc}
\end{defn}

We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
some initial memory fixed throughout our development, mapping every
address to value $v_0$; and $\theta_0$ maps every processor ID to
$(\pc_0, s_0)$, using architecture-specific default values $s_0$ and
$\pc_0$.

This LTS encodes Lamport's notion of SC, where processors take turns executing
nondeterministically in a simple interleaving.  We need know very few details
of the ISA to define the SC model.  Our final, optimized system is parametrized
over an ISA in the same way.  In the course of the rest of this paper, we will
define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.

Note that, in this setting, an operational specification like the LTS
for SC is precisely the right characterization of \textbf{full functional
  correctness} for a hardware design, much as a
precondition-postcondition pair does that job in a partial-correctness
Hoare logic.  Our SC LTS fully constrains observable behavior of a
system to remain consistent with simple interleaving.  Similar
operational models are possible as top-level specifications for
systems following weaker memory models, by giving the LTS for the \emph{Local
Buffer} component and composing the three components simultaneously.

\section{Speculative Out-Of-Order Processor}\label{sec:ooo}

The previous section divided a multiprocessor system into two rather
simple components, processors and memories.  We are now ready to begin
developing more complex, optimized versions of each piece, starting
with processors.

We implement a \emph{speculative} processor, which may create many
simultaneous outstanding requests to the memory, as an optimization to
increase parallelism.  Our processor proof is in some sense generic
over correct speculation strategies.  We parametrize over two key
components of a processor design: a \emph{branch predictor}, which
makes guesses about processor-local control flow in advance of
%receiving enough memory responses to resolve
resolving conditional jumps; and a
\emph{reorder buffer}, which decides what speculative instructions (like memory loads)
are worth issuing at which moments (in effect \emph{reordering} later
instructions to happen before earlier instructions have finished).

The branch predictor is the simpler of the two components to model.
We indicate branch-predictor state with metavariable $\ppc$.
The operations on such state are $\curPpc(\ppc)$, to extract the
current program-counter prediction; $\nextPpc(\ppc)$, to ask the
predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
to reset prediction to begin at a known-accurate position $\pc$.  It
turns out that we need to impose no explicit correctness criterion on
branch predictors.  The processor uses predictions only as hints, and
it always resets the predictor using $\setBp$ after detecting an
inaccurate hint.

The interface and formal contract of a reorder buffer are more
involved.  We write $\rob$ as a metavariable for reorder buffer
state.  The associated operations are:
\begin{itemize}
\item $\phi$, the state of an empty buffer.
\item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
\item $\get\spc\ld(\rob)$, which returns a set of speculative loads that can be issued to the memory system
\item $\compute(\rob)$, which models a
step of computation inside the buffer. For instance, it invokes the $\dec$
and $\exec$ functions (as defined for SC) internally to obtain the next program
counter, state, \etc{} (only the speculative states are updated, not the actual states).
\item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
has returned result value $v$ for the speculative load with tag $t$.
\item $\commit(\rob)$, which returns the next instruction in serial
program order, if we have accumulated enough memory responses to execute it
accurately, or returns
$\epsilon$ otherwise.  When $\commit$ returns an instruction, it also
returns the associated program counter plus the next program counter
we would advance to afterward.
 Furthermore, the instruction is
extended with any relevant response from memory (used only for load
instructions, obtained through $\upd\ld$) and with the new architectural state (\eg{} register
file) after execution.
\item $\retire(\rob)$, which informs the buffer that its $\commit$
instruction was executed successfully, so it is time to move on to the
next instruction.
\end{itemize}

We postpone characterization of correct reorder buffers until after we
present the LTS for speculating processors.

\begin{figure*}[t]
\small
\centering
\begin{boxedminipage}[c]{.95\textwidth}
\inference[Fetch]
{}
{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}

\inference[Compute]
{}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}

\inference[SpecLd]
{(\spc\ld,t,a) \in \get\spc\ld(\rob)}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\rob,t,v),\ppc}
{(\ld\req, a), (\ld\resp, v), \Spec}}

\inference[Abort]
{\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\phi,\setBp(\ppc,\pc)}}

\inference[Halt]
{\commit(\rob) = H}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{H}}

\inference[Nm]
{\commit(\rob) = (\pc,\pc',(\nm,s'))}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}}

\inference[St]
{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
{\io{\text{P$_\text{so}$}}{s',\pc',\retire(\rob),\ppc}{s,\pc,\rob,\ppc}
{\st, a, v}}

\inference[LdGood]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v), \Real}}

\inference[LdBad]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'', \pc'')}
{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v), \Real}}
\end{boxedminipage}

\caption{Speculating, out-of-order issue processor}
\label{Ooo}
\end{figure*}

Figure~\ref{Ooo} defines that LTS,
P$_{\texttt{so}}$. This processor may issue arbitrary speculative
loads, but it only ever \emph{commits} the next instruction in serial
program order. 
We will see the processor issue two kinds of loads, a speculative load (with
the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
maintain SC, every speculative
load must have a matching verification load later on, and we maintain
the illusion that the program only depends on the results of
verification loads, which, along with stores, \emph{must be issued in
serial program order}.

When committing a
previously issued speculative load instruction, the associated speculative
memory load response is verified against the new commit load response. If the
resulting values do not match, the processor terminates all past uncommitted
speculation, by emptying the reorder buffer and resetting the
next predicted program counter in the branch predictor to the correct next value. This step mimics
the mechanism implemented in most modern-day processors, operating on the
conservative assumption that the processor may have performed arbitrary
incorrect speculation after receiving the inaccurate response from
memory.  (Note that here ``inaccurate'' does not mean ``wrong at
the time the response was received,'' but rather ``wrong at commit
time'' or ``correct before, but not anymore.'')

A full processor state is $(s, \pc, \rob, \ppc)$, comprising
architectural state, the program counter, and the reorder buffer and branch
predictor states. Its initial state is given by $(s_0, \pc_0, \phi, \ppc_0)$.
The interface of this processor with memory is identical
to that of the reference processor, but now we can exercise a memory
system's transitions even for speculative loads.

\begin{figure*}[t]
\small
\begin{inv}
If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}

$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\phi,\ppc_0}{s,\pc,\rob,\ppc}{\eta} \Rightarrow$
\begin{displaymath}
\left\{
\begin{array}{ll}
\commit(\rob) = (\pc, \pc', (\nm, s')) &\Rightarrow
\exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
(s', \pc')\\
\commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (s',\pc')\\
\commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
\dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
(s',\pc')\\
\commit(\rob) = H & \Rightarrow
\dec(s, \pc) = H\\
\end{array}
\right.\end{displaymath}
\label{rob}
\end{inv}
\vspace{-.5cm}
\caption{Correctness of reorder buffer}
\label{robfig}
\end{figure*}

Finally, we impose a general correctness condition on reorder
buffers.  They must maintain Invariant~\ref{rob} in Figure~\ref{robfig}.
Intuitively, whenever the buffer claims in a $\commit$
output that a particular instruction is next to execute, causing
certain state changes, that instruction must really be next in line according
to how the program runs in the SC system, and running it must really cause
those state changes.

When this condition holds, we may conclude the correctness theorem for
out-of-order processors.  We use a trace-transformation function
$\noSpec$ that drops all speculative-load requests and responses.
See Definition~\ref{refines} for a review of how we use such
functions in framing trace refinement.  Intuitively, we prove that any
behavior by the speculating processor can be matched by the simple
processor, with speculative messages erased.

\begin{theorem}
\label{ocorrect}
$\text{P$_\text{so}$} \sqsubseteq_\noSpec$ P$_\text{ref}$
\end{theorem}
\begin{proof}
By induction on P$_\text{so}$ traces, using a proper abstraction
function (which drops the speculative messages and the $\rob$ and $\ppc$ states) to relate the two systems.  Invariant~\ref{rob} is crucial to
relate the reorder buffer's behavior with the simple in-order
execution of P$_\text{ref}$.
\end{proof}

\begin{corollary}
\label{ges}
$\text{P$^n_\text{so}$} \sqsubseteq_{\noSpec^n}$ P$_\text{ref}^n$
\end{corollary}
\begin{proof}
Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
latter is one of our preliminary results about $n$-repetition).
\end{proof}

@@

We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
denote that $c$ is a leaf or L1 cache.

@@

%\medskip

@@

%Dropping the request
%removes the entry from the buffer, allowing the child to process further messages
%from the parent -- otherwise there would have been a deadlock.

\begin{figure*}
\small
\centering

\begin{boxedminipage}[c]{.95\textwidth}

\textbf{Processor/Memory Interface}

\inference[Load]
{\s(c,a) \geq S}
{\io{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v), \_)}}
\vspace{.1in}
\inference[Store]
{s(c,a) \geq M}
{\io{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}

\textbf{Child Upgrade}

\inference[ChildSendReq]
{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
\wt[(c,a)\coloneqq x], \dwt}}
\vspace{.1in}
\inference[ParentRecvReq]
{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
\wt, \dwt}}
\vspace{.1in}
\inference[ChildRecvRsp]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}

\textbf{Parent Downgrade}

\inference[ParentSendReq]
{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
\wt, \dwt[(p, c,a)\coloneqq x]}}
\vspace{.1in}
\inference[ChildRecvReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt}}
\vspace{.1in}
\inference[ParentRecvRsp]
{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt}
{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}

\textbf{Voluntary downgrade for replacement}

\inference[VolResp]
{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt}
{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt}}

\textbf{Dropping request because of voluntary downgrade}

\inference[DropReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
 \s(c,a) \le x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt}
{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
\wt, \dwt}}
\end{boxedminipage}

\caption{LTS for cache-coherent shared-memory system}
\label{cache}
\end{figure*}

@@

We first reduce the proof of store atomicity to several lemmas. We then
present a key insight that we discovered for proving these lemmas in the form
of three fundamental invariants. We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

>>>>

@@

We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
$\wt$, $\dwt$, $\inp$, $\outp$. The $\inp$ and $\outp$ sets are the interfaces
to the processors and are exactly the same as in the simple memory LTS
(Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.

%We describe a state of the system using 6 fields: $\dt$, $\ch$, $\s$, $\dst$,
%$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
denote that $c$ is a leaf or L1 cache.

@@

@@

\input{TransitionsUsefulCache1}

@@

We first reduce the proof of store atomicity to several lemmas.
We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

The lemmas
presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each lemma, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.

Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.
